Floral preferences of mountain bumble bees are constrained by trait matching but flexible through elevation and season

### 2021-09-02 ms_2/ms_2_oikos_part2.Rmd
beginning ms1 revision

---
title: "Floral preferences of mountain bumble bees are constrained by trait matching but flexible through elevation and season"
header-includes:
   - \usepackage{lineno}\linenumbers
   - \usepackage{setspace}\doublespacing
output: pdf_document
bibliography: 
  - references.bib
csl: ecologyletters.csl
---

**Douglas Sponsler** (corresponding author)  
University of Würzburg  
Department of Animal Ecology and Tropical Biology  
Biozentrum der Universität Würzburg C-018  
Am Hubland  
97074 Würzburg, Germany  
Email: douglas.sponsler@uni-wuerzburg.de  
Telephone: +49 0176 363 20273

**Katharina Kallnik**  
University of Würzburg  
Department of Animal Ecology and Tropical Biology  
97074 Würzburg, Germany  
Email: katharina.kallnik@uni-wuerzburg.de

**Fabrice Requier**  
Université Paris-Saclay  
Évolution, Génomes, Comportement et Écologie  
91198 Gif-sur-Yvette, France  
Email: fabrice.requier@egce.cnrs-gif.fr

**Alice Classen**  
University of Würzburg  
Department of Animal Ecology and Tropical Biology  
97074 Würzburg, Germany  
Email: alice.classen@uni-wuerzburg.de

**A. Fabienne Maihoff**  
University of Würzburg  
Department of Animal Ecology and Tropical Biology  
97074 Würzburg, Germany  
Email: anne_fabienne.maihoff@uni-wuerzburg.de

**Johanna Sieger**  
Naturpark Frankenhöhe  
91522 Ansbach, Germany   
Email: sieger@lpv-mfr.de

**Ingolf Steffan-Dewenter**  
University of Würzburg  
Department of Animal Ecology and Tropical Biology  
97074 Würzburg, Germany  
Email: ingolf.steffan@uni-wuerzburg.de

**AUTHORSHIP STATEMENT:** KK and ISD conceived the study. KK, JS, and ISD designed and carried out field work. DS analyzed the data and wrote the paper with substantial contributions from FR, AC, AFM, KK, and ISD.

**DATA ACCESSIBILITY:** All data used in this study have been archived on DRYAD (https://datadryad.org/stash/dataset/doi:10.5061/dryad.s1rn8pk7f) and will be available upon article publication.

**ARTICLE TYPE:** Letter

**RUNNING TITLE:** Bumble bee floral preferences

**KEYWORDS:** foraging, trait matching, pollinator, *Bombus*, alpine, niche, resource selection

**SECTION LENGTHS:**  
Abstract:  
Main text:  
References:  
Figures:

@@

Patterns of resource use by animals can clarify how ecological communities have assembled in the past, how they currently function, and how they are likely to respond to future perturbations. Using ~13,000 records of bumble- bee-wildflower interactions along a 1400 m elevation gradient, we find that the selection of floral morphotypes by bumble bees is driven both by the directed process of trait-matching and the neutral effect of floral abundance. Patterns of floral preference also vary by elevation and season, which may be an adaptation to the strong spatiotemporal turnover of mountain floral communities. The flexibility of bumble bee foraging will be tested in coming years as climate and land use change drive rapid changes in floral resource distribution in time and space, particularly in the mountain ecosystems that host the world's richest communities of bumble bees and their floral mutualists.

@@

<!-- Plant-pollinator networks are a special case of resource selection in which consumer fitness and resources fitness are positively correlated, causing the traits of plants and pollinators to converge toward compatibility with one another [e.g. @Newman2014-ke]. The coexistence of multiple plant and pollinator species, however, requires niche differentiation [@Palmer2003-ih; @Johnson2019-su], which can be achieved by trait-based partitioning of interactions among pollinators and their floral hosts [@Willmer2011-ak]. -->

@@

Addressing the question of how bumble bees will select floral resources under novel conditions requires discernment of the generalizable mechanisms that underpin patterns of floral selection observed under current conditions. It is ordinarily assumed that the choice of resources by foraging animals is a joint function of trait-based preferences and the neutral effects of the relative abundances of available resources: the so-called "use-availability" model of resource selection [e.g. @Beyer2010-vr]. The most well-studied mechanism of the preference component of bumble bees floral selection is trait-matching between bumble bee tongue length and floral corolla depth. Bumble bee species can differ markedly in tongue length, and floral handling time increases when tongue length and corolla depth are poorly matched [@Inouye1980-mv; @Harder1983-bk], leading to an association between tongue length and floral preferences in bumble bee communities [e.g. @Brian1954-cy; @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn]. Under the use-availability model, tongue-length-based floral preferences should be modulated by the relative abundance of available resources to generate observed patterns of floral selection. It has recently been proposed, however --- in what has been termed the *target use hypothesis* --- that floral selection by bumble bees is a process of *pure preference*, irrespective of resource abundance [@De_Valpine2013-be; @Harmon-Threatt2017-bz]. If and when the *target use hypothesis* holds, observed visitation rates are a better index of preference than metrics designed to control for resource abundance (e.g. null model residuals), since the subtraction of presumed neutral effects of abundance when those effects do not really exist actually creates new abundance-dependent artifacts [@Harmon-Threatt2017-bz]. On the other hand, failing to account for neutral effects of abundance when they do exist similarly obscures the general mechanisms behind observed foraging patterns [@Vazquez2003-qa; @Bluthgen2008-rp] that will determine bumble bee resource selection under novel conditions induced by climate and land use change.

@@

Field sampling was conducted 2010-2012 in Berchtesgaden National Park (47.55°N, 12.92°E), located in the Northern Limestone Alps of southeast Germany. The landscape is composed of a coniferous forest matrix embedded with pastures either currently or historically maintained by extensive (i.e. low intensity) agricultural management. We selected a total of 25 study sites (60 x 60 m) located on pastures along a ~1400 m elevation gradient ranging from 641-2032 m above sea level (m.s.l.). Fourteen of these pastures are grazed by cattle or sheep, three are mowed for hay production, and eight have been abandoned throughout the last century and are no longer subject to human management.

@@

Bumble- bee-wildflower interactions were observed during repeated visits (hereafter, “visitation samples”) to each study site at approximately weekly intervals. Sampling was only conducted during periods without rain and with an air temperature of at least 6°C. Bumble bees were recorded during one 50-minute transect walk per visit, and bumblebees observed on a given flower were counted as floral visitors. Queens were identified to species level in the field to avoid destructively sampling of colony foundresses, but workers and males were collected and identified in the lab following @Amiet1996-jz. For analysis, pooled the cryptic *Bombus lucorum/terrestris* species complex and the parasitic *Bombus* (*Psithyrus*) species, respectively, into species-groups. Males were recorded, we analyzed only the pooled visitation of queens and workers. In conjunction with each bumble bee visitation sample, we estimated the flower cover of herbaceous and shrubby plant species within each 60 x 60 m study plot to the nearest 0.01 m^2^ (hereafter "floral survey samples") and identified the plant species according to @Lauber2007-ha and @Oberndorfer2001-ro. See Sponsler et al. (submitted) for details of sampling design.

@@

To summarize floral morphology, we relied on the synthesis of @Kugler1970-gb, who developed a typology consisting of ten primary morphotypes and nested subdivision of each. Kugler’s typology was informed by laboratory studies of bumble bee flower choice [@Kugler1943-uy], and there is a precedent for analyzing bumble bee visitation in terms of Kugler’s morphotypes [@Farnberger1996-xo; @Neumayer1999-il; @Schneller2014-ry]. Kugler morphotype classifications of the plants in our study system were accessed via the BIOFLOR database [@Klotz2002-uc]. Morphotypes were simplified to the primary classes numbered 0-10: 0 = null (includes wind-pollinated flora such as grasses), 1 = disc- and bowl-shaped flowers, 2 = funnel flowers, 3 = bell-shaped flowers, 4 = stalk-disc flowers, 5 = lip flowers, 6 = flag flowers, 7 = flower heads, 8 = spike flowers, 9 = brush flowers, 10 = trap flowers **(Figure 1)**. In a few cases, a plant species in our data set was not found in the BIOFLOR database. In most cases, morphotypes are constant at the genus level, and missing morphotypes were added based on the morphotypes found for congeners. In the case of the genus *Phyteuma*, we followed @Neumayer1999-il and classified it as a group 7 flower head. Morphotypes 9 and 10 were extremely rare in our study system and morphotype 8 was completely absent, so these morphotypes were omitted from all analyses. Plants belonging to morphotype 0 were typically not recorded during floral surveys due to their lack of conspicuous floral displays, and they were rarely visited by bumble bees, so this morphotype was also omitted from all analyses.

@@

For the purposes of our study, we define "preference" as the difference between the observed visitation rate to a given resource and visitation rate predicted by the relative abundance of that resource (preference can thus take negative values for under-visited resources). For more nuanced definitions of preference, we refer the reader to @Beyer2010-vr and references therein.

@@

To quantify preference, we compared the proportional visitation of each morphotype (by each bumble bee species) with the rates of visitation expected under a null model constructed following @Vaughan2018-gj. In this modeling approach, resource selection is simulated for individual consumers --- in our case, individual bumble bees --- according to a resource selection process in which the probability of a resource being chosen is proportional to its relative abundance [@Agusti2003-tt]. By simulating the visitation of individual consumers rather than species in aggregate, the abundance of each resource, and therefore its probability of visitation, can vary across site-dates, thus preserving the within-sample correspondence between visitation and abundance. Simulated interactions are then summarized for each consumer species by taking the mean number of interactions with each resource across all individuals of the consumer species and all samples of the dataset. As this process is iterated, the result is a distribution of visitation rates of each consumer species to each resource under the null assumption that resource selection is determined only by resource abundance. The visitation distributions of the null model can then be compared to observed visitation rates to infer positive or negative preference of each consumer toward each resource. Since the mean of the null distribution can be interpreted as the expected value under null assumptions, the difference between the mean null value and the observed value can be interpreted as a magnitude of preference [@Vaughan2018-gj], and statistical significance can be estimated as the proportion of simulated null values more extreme than the observed value.

@@

We followed the procedure described above, simulating our entire dataset in one modeling process to infer aggregate preference between bumble bee species and floral morphotypes. Using the results of this analysis, we modeled preference toward each floral morphotype as a function of bumble bee tongue length using hierarchical generalized additive modeling (HGAM) [@Pedersen2019-kf]. As a test of the *target use hypothesis*, we repeated the HGAM analyses described above using observed visitation rates rather than preference. To verify that our inference of preference was not sensitive to our decision to pool queen and worker visitation, we recalculated the null model and tongue length HGAM using only data from workers. Finally, we compared morphotype preference across bumble bee species by performing a principal components analysis on the bumble bee species x floral morphotype preference matrix, and we mapped bumble bee tongue length onto the ordination space by fitting a 2-dimensional GAM surface to the first two principal components.

@@

To explore temporal and elevational variation in floral preference, we repeated the null model analysis described above but this time for each site-date individually, so that the preference of each bumble bee species toward each floral morphotype could be modeled as a variable through elevation and time. For this analysis, we included only the six most abundant and widespread bumble bee species in our study system (*B. hortorum*, *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris-lucorum*, and *B. wurflenii*) to ensure sufficient data points through elevation and time, and we modeled their preference only toward the 4 most visited floral morphotypes (bell, lip, flag, and head flowers). Preference for each bumble- bee-morphotype pair (hereafter "bb-fl pair") was modeled together in a single HGAM with an interaction smooth between elevation and time.

@@

An inherent problem with modeling sample-wise variation in preference is that the bounds of preference are strictly dependent on floral abundance [@Beyer2010-vr]. A morphotype present at very low relative abundance would, under the null model, be expected to receive very low visitation, which sets the lower bound of preference near zero (i.e. neutrality) but allows the upper bound to approach 1. Conversely, a morphotype present at very high relative abundance would receive very high visitation under the null model, and its upper bound of preference would fall near zero and its lower bound near -1. This clearly creates a degree of unavoidable confoundedness between preference and floral abundance. With this limitation in mind, we fit HGAMs of the same form to the relative abundance of each floral morphotype, and we present the results together to be interpreted jointly.

@@

All data processing and analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. Null models were constructed using `econullnetr` [@Vaughan2018-gj]. GAM analyses were performed with `mgcv` [@Wood_2017] and visualized with `mgcViz` [@Fasiolo2018aa] and `tidymv` [@Coretta2021-aa]. PCA was performed with the `vegan` [@Oksanen_2019] and visualized using `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Annotated R code is available in the Appendix X.

@@

Preference was generally strong, with the visitation rates of most bumblebee species to most floral morphotypes falling significantly above or below the 99% interval of the null model **(Figure 1)**. Nonsignificant preference was observed mainly among the rarer bumblebee species, likely because their small samples sizes lead to wide null model confidence intervals. Interestingly, bumble bees showed strong species-specific variation in their relative preferences of different floral morphotypes **(Figure 1)**. The omission of queens from the calculation of overall preference did not yield qualitatively different results (Appendix X).

@@

When floral morphotypes are compared to one another in terms of preference across all bumble bee species, clear patterns emerge **(Figure 2A)**. All species exhibited negative preference toward disc flowers, while most species exhibited positive preference toward bell, lip, and flag flowers. preference toward funnel and stalk-disc flowers tended to be neutral (with one strong positive outlier --- *B. gerstaeckeri* --- for funnel flowers), reflecting the fact that these morphotypes tended both to be rarely visited and to occur at low abundance. Preference toward head flowers varied widely across bumble bee species, ranging from strongly positive to strongly negative.

@@

<!-- I could do some sort of linear mixed effects model on this I suppose, but I think it would violate assumptions due to the non-independence of preference (preference for one thing implies preference against others). I've already sort of done stats on this with the null model. -->

@@

Tongue length was a strong predictor of preference but not of observed visitation rates **(Figure 3)**. Preference toward funnel, bell, lip, flag, and head flowers exhibited significant GAM curves in response to bumble bee tongue length, and the model explained 77% of total deviance (adj. R^2^ = 0.728). For funnel and lip flowers, preference increased linearly with tongue length. Preference toward bell flowers and head flowers exhibited approximately quadratic responses with preference peaking among short-tongued species and declining sharply with increasing tongue length before leveling off. Preference toward flag flowers exhibited a unimodal peak among medium-tongued species. In contrast, observed visitation rates were predicted by tongue length only for head flowers, which exhibited a weak decline in preference from a peak among short-tongued species. The visitation rate model explained only 25% of total deviance (adj. R^2^ = 0.053). Again, these results did not differ when preference was based on the worker-only null model (Appendix X).

@@

The proportional abundance of the most visited floral morphotypes --- bell, lip, flag, and head flowers --- was strongly patterned by elevation and season **(Figure 5)**. Bell flowers peaked in proportional abundance early in the season and at mid-elevation, and lip flowers peaked late in the season and at low elevation. Flag flowers peaked in mid-season and exhibited two distinct elevational peaks, the lower at around 1000 m.s.l and the higher at around 1800 m.s.l. Head flowers exhibited a mid-season peak at low elevation and a broad late season peak at around 1600 m.s.l. All smooths were significant (p << 0.0001), and the model explained 64.3% of total deviance (adj. R^2^ = 0.554).

@@

The morphotype preferences of bumble bee species were also strongly patterned through elevation and season. In some cases, though, patterns of preference mirrored patterns of floral availability, suggesting that bumble bees may exhibit functional responses [@Mysterud1998-rs] to floral abundance not entirely accounted for by null models that assume a linear proportionality between abundance and visitation. Preference for flag flowers was strongly associated with high elevation, except among the species that tended to avoid flag flowers (*B. pratorum* and *B. soroeensis*) **(Figure 1)**. It is also apparent that, as a rule, preference is more dynamic where it is also higher overall. Bell flowers, which were positively selected overall by the short-tongued bumble bee species *B. pratorum*, "*B. terrestris-lucorum*, and *B. soroeensis* **(Figure 1)**, also exhibited the most complex patterns of preference in elevation and time for these bumble bee species **(Figure 6)**. In contrast, the long and medium-tongued species *B. hortorum*, *B. pascuorum*, and *B. wurflenii*, which exhibited neutral or weakly positive overall selection toward bell flowers **(Figure 1)**, also show weak patterns of variation in preference toward bell flowers **(Figure 6)**. Nearly the reverse pattern is evident for flag flowers, reflecting their association with longer-tongued bumble bee species.

@@

Smooths were significant (p < 0.05) for all bee-morphotype pairs except hort-head, prat-flag, telu-head, and soro-flag. The model explained 36.4% of total deviance (adj. R^2^ = 0.33).

@@

Using ~13,000 bumble- bee-wildflower interactions, we characterized bumble bee floral selection in terms of generalized floral morphotypes and found evidence that bumble bee floral selection depends both on the directed process of morphotype preference and the neutral effect of floral abundance. Bumble bee species differed in their floral morphotype preferences, and these differences could be explained by trait-matching between floral morphotype and bumble bee tongue length. These preferences, however, varied by elevation and season, likely representing adaptive responses of bumble bee foraging behavior to elevational and seasonal shifts in floral resource composition, possibly modulated by competitive interactions among bumble bees species with overlapping resource preferences.

@@

Our finding that bumble bee tongue length explained preference (i.e. null model residuals) much better than it explained observed visitation rates indicates that, in our study system, abundance and preference function jointly in bumble bee floral selection. This casts doubt on the *target use hypothesis* as a general rule of bumble bee foraging, though it is possible that it still obtains under certain circumstances [e.g. @Harmon-Threatt2017-bz]. In general, though, our findings support the hypothesis that bumble bee resource selection involves both the neutral effects of floral abundance and the directed effect of preference, meaning that analytical techniques designed to disentangle these effects, such as the null model method we employ, should be favored in studies of bumble bee foraging.

@@

We found that the most preferred morphotypes were bell, lip, and flag flowers, which is broadly consistent with visitation patterns reported by previous studies in central Europe [@Schneller2014-ry; @Neumayer1999-il; Farnberger1996-xo], indicating that our findings are regionally generalizeable. These previous studies tended to find a stronger role of head flowers than indicated by our preference analysis, but this is probably because they characterized foraging in terms of simple visitation rates rather than null-model-inferred preference. We saw similarly high visitation rates toward head flowers, but our null-model analysis reveals that while head flowers were highly visited, they tended to be under-visited relative to their abundance.

@@

**Figure 1:** Observed visitation rates compared to null model simulations for each bumble bee species and floral morphotype (A). Gray bars depict 99% intervals of null model simulations and points depict observed visitation rates. Points are color-coded according to where they fall with respect to the 99% interval of the null model: red = above (positive preference), blue = below (negative preference), gray = within (nonsignificant preference). The difference between the observed value and the center of the null model interval can be interpreted as magnitude of preference. Each floral morphotype is illustrated (B) with a representative species common in our study system: 1 = *Potentilla erecta*, 2 = *Gentiana asclepiadea*, 3 = *Campanula glomerata*, 4 = *Silene nutans*, 5 = *Ajuga reptans*, 6 = *Lotus corniculatus*, 7 = *Centaurea jacea*. Illustrations are modifications of public domain works obtained from www.plantillustrations.org. For details of original sources, see Appendix X.

>>>>

---
title: "Floral preferences of mountain bumble bees are constrained by trait matching but flexible through elevation and season"
header-includes:
   - \usepackage{lineno}\linenumbers
   - \usepackage{setspace}\doublespacing
output: 
  pdf_document: 
    keep_tex: yes
bibliography: 
  - references.bib
csl: oikos.csl
---

@@

Patterns of resource use by animals can clarify how ecological communities have assembled in the past, how they currently function, and how they are likely to respond to future perturbations. Using ~13,000 records of bumble- bee-wildflower interactions along a 1400 m elevation gradient, we find that the selection of floral morphotypes by bumble bees is driven both by the directed process of trait-matching and the neutral effect of floral abundance. Patterns of floral preference also vary by elevation and season, which may be an adaptation to the strong spatiotemporal turnover of mountain floral communities. The flexibility of bumble bee foraging will be tested in coming years as climate and land use change alter floral resource distribution in time and space, particularly in the mountain ecosystems that host the world's richest communities of bumble bees and their floral mutualists.

**KEYWORDS:** foraging, trait matching, pollinator, *Bombus*, niche, resource selection

@@

@@

Addressing the question of how bumble bees will select floral resources under novel conditions requires discernment of the generalizable mechanisms that underpin patterns of floral selection observed under current conditions. It is ordinarily assumed that the choice of resources by foraging animals is a joint function of trait-based preferences and the neutral effects of the relative abundances of available resources: the so-called "use-availability" model of resource selection [e.g. @Beyer2010-vr]. The most well-studied mechanism of the preference component of bumble bee floral selection is trait-matching between bumble bee tongue length and floral corolla depth. Bumble bee species can differ markedly in tongue length, and floral handling time increases when tongue length and corolla depth are poorly matched [@Inouye1980-mv; @Harder1983-bk], leading to an association between tongue length and floral choice in bumble bee communities [e.g. @Brian1954-cy; @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn]. Under the use-availability model, tongue-length-based floral preferences should be modulated by the relative abundance of available resources to generate observed patterns of floral selection. It has recently been proposed, however --- in what has been termed the *target use hypothesis* --- that floral selection by bumble bees is a process of *pure preference*, irrespective of resource abundance [@De_Valpine2013-be; @Harmon-Threatt2017-bz]. If and when the *target use hypothesis* holds, observed visitation rates are a better index of preference than metrics designed to control for resource abundance (e.g. null model residuals), since the subtraction of presumed neutral effects of abundance when those effects do not really exist actually creates new abundance-dependent artifacts [@Harmon-Threatt2017-bz]. On the other hand, failing to account for neutral effects of abundance when they do exist similarly obscures the general mechanisms behind observed foraging patterns [@Vazquez2003-qa; @Bluthgen2008-rp] that will determine bumble bee resource selection under novel conditions induced by climate and land use change.

@@

Field sampling was conducted 2010-2012 in Berchtesgaden National Park (47.55°N, 12.92°E), located in the Northern Limestone Alps of southeast Germany. The landscape is composed of a coniferous forest matrix embedded with pastures either currently or historically maintained by extensive (i.e. low intensity) agricultural management. We selected a total of 25 study sites (60 x 60 m) located on pastures along a ~1400 m elevation gradient ranging from 641-2032 m above sea level (m.s.l.). Fourteen of these pastures are grazed by cattle or sheep, three are mowed for hay production, and eight have been abandoned during the last century and are no longer subject to human management.

@@

Bumble- bee-wildflower interactions were observed during repeated visits to each study site at approximately weekly intervals. Sampling was only conducted during periods without rain and with an air temperature of at least 6°C. Bumble bees were recorded during one 50-minute transect walk per visit, and bumblebees observed on a given flower were counted as floral visitors. Queens were identified to species level in the field to avoid destructively sampling colony foundresses, but workers and males were collected and identified in the lab following @Amiet1996-jz. Prior to analysis, we pooled the cryptic *Bombus lucorum/terrestris* species complex and the parasitic *Bombus* (*Psithyrus*) species, respectively, into species-groups. Males were recorded, but we analyzed only the pooled visitation of queens and workers. In conjunction with each bumble bee visitation sample, we estimated the flower cover of herbaceous and shrubby plant species within each 60 x 60 m study plot to the nearest 0.01 m^2^ and identified the plant species according to @Lauber2007-ha and @Oberndorfer2001-ro. See Sponsler et al. (submitted) for details of sampling design.

@@

To summarize floral morphology, we relied on the synthesis of @Kugler1970-gb, who developed a typology consisting of ten primary morphotypes and nested subdivisions of each. Kugler’s typology was informed by laboratory studies of bumble bee flower choice [@Kugler1943-uy], and there is a precedent for analyzing bumble bee visitation in terms of Kugler’s morphotypes [@Farnberger1996-xo; @Neumayer1999-il; @Schneller2014-ry]. Kugler morphotype classifications of the plants in our study system were accessed via the BIOFLOR database [@Klotz2002-uc]. Morphotypes were simplified to the primary classes numbered 0-10: 0 = null (includes wind-pollinated flora such as grasses), 1 = disc- and bowl-shaped flowers, 2 = funnel flowers, 3 = bell-shaped flowers, 4 = stalk-disc flowers, 5 = lip flowers, 6 = flag flowers, 7 = flower heads, 8 = spike flowers, 9 = brush flowers, 10 = trap flowers **(Figure 1)**. In a few cases, a plant species in our data set was not found in the BIOFLOR database. In most cases, morphotypes are constant at the genus level, and missing morphotypes were added based on the morphotypes found for congeners. In the case of the genus *Phyteuma*, we followed @Neumayer1999-il and classified it as a group 7 flower head. Morphotypes 9 and 10 were extremely rare in our study system and morphotype 8 was completely absent, so these morphotypes were omitted from all analyses. Plants belonging to morphotype 0 were typically not recorded during floral surveys due to their lack of conspicuous floral displays, and they were rarely visited by bumble bees, so this morphotype was also omitted from all analyses.

@@

For the purposes of our study, we define "preference" as the difference between the observed visitation rate to a given resource and visitation rate predicted by a null model in which the probability of a resource being chosen is proportional to its relative abundance [@Agusti2003-tt]. In other words, preferences are null model residuals, and thus take negative values for under-visited resources and positive values for over-visited resources. For more nuanced definitions of preference, we refer the reader to @Beyer2010-vr and references therein.

@@

We constructed null models following @Vaughan2018-gj. In this modeling approach, resource selection is simulated for individual consumers --- in our case, individual bumble bees. By simulating the visitation of individual consumers rather than species in aggregate, the abundance of each resource, and therefore its probability of visitation, can vary across samples (i.e. site-dates), thus preserving the within-sample correspondence between visitation and abundance. Simulated interactions are then summarized for each consumer species by taking the mean number of interactions with each resource across all individuals of the consumer species and all samples of the dataset. As this process is iterated, the result is a distribution of visitation rates of each consumer species to each resource under the null assumption that resource selection is determined only by resource abundance. The visitation distributions of the null model can then be compared to observed visitation rates to infer positive or negative preference of each consumer toward each resource. Since the mean of the null distribution can be interpreted as the expected value under null assumptions, the difference between the mean null value and the observed value can be interpreted as a magnitude of preference [@Vaughan2018-gj], and statistical significance can be estimated as the proportion of simulated null values more extreme than the observed value.

@@

We followed the procedure described above, simulating our entire dataset in one modeling process to infer aggregate preferences of bumble bee species for each floral morphotype. Using the results of this analysis, we modeled preferences for each floral morphotype as a function of bumble bee tongue length with hierarchical generalized additive modeling (HGAM), following the *type I* effect structure described by @Pedersen2019-kf. As a test of the *target use hypothesis*, we repeated the HGAM analyses described above using observed visitation rates rather than preference. To verify that our inference of preference was not sensitive to our decision to pool queen and worker visitation, we recalculated the null model and tongue length HGAM using only data from workers. Finally, we compared morphotype preferences across bumble bee species by performing a principal components analysis on the bumble bee species x floral morphotype preference matrix, and we mapped bumble bee tongue length onto the ordination space by fitting a 2-dimensional GAM surface to the first two principal components.

@@

To explore temporal and elevational variation in floral preference, we repeated the null model analysis described above but this time for each site-date individually, so that the preference of each bumble bee species toward each floral morphotype could be modeled as a variable through elevation and time. For this analysis, we included only the six most abundant and widespread bumble bee species in our study system (*B. hortorum*, *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris-lucorum*, and *B. wurflenii*) to ensure sufficient data points through elevation and time, and we modeled their preference only toward the 4 most visited floral morphotypes (bell, lip, flag, and head flowers). Preference for each bumble- bee-morphotype pair was modeled together in a single HGAM with an interaction smooth between elevation and time. We also included bee-morphotype pair as a fixed itnercept effect and the interactions between bee-morphotype pair and site and bee-morphotype pair and year as random effects.

@@

An inherent problem with modeling sample-wise variation in preference is that the bounds of preference are strictly dependent on floral abundance [@Beyer2010-vr]. A morphotype present at very low relative abundance would, under the null model, be expected to receive very low visitation, which sets the lower bound of preference near zero (i.e. neutrality) but allows the upper bound to approach 1. Conversely, a morphotype present at very high relative abundance would receive very high visitation under the null model, and its upper bound of preference would fall near zero and its lower bound near -1. This creates some confoundedness between preference and floral abundance. With this limitation in mind, we fit HGAMs of the same form to the relative abundance of each floral morphotype, and we present the results together to be interpreted jointly.

@@

All data processing and analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. Null models were constructed using `econullnetr` [@Vaughan2018-gj]. GAM analyses were performed with `mgcv` [@Wood_2017] and visualized with `mgcViz` [@Fasiolo2018aa] and `tidymv` [@Coretta2021-aa]. PCA was performed with the `vegan` [@Oksanen_2019] and visualized using `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Annotated R code and complete output of our analytical workflow is available in Appendix 1.

@@

Preferences were generally strong, with the visitation rates of most bumblebee species to most floral morphotypes falling significantly above or below the 99% interval of the null model **(Figure 1)**. Nonsignificant preference was observed mainly among the rarer bumblebee species, likely because their small samples sizes lead to wide null model confidence intervals. Notably, bumble bees showed strong species-specific variation in their relative preferences of different floral morphotypes **(Figure 1)**. The omission of queens from the calculation of overall preference did not yield qualitatively different results (Appendix 1).

@@

When floral morphotypes are compared to one another in terms of preference across all bumble bee species, clear patterns emerge **(Figure 2A)**. All species exhibited negative preference toward disc flowers, while most species exhibited positive preference toward bell, lip, and flag flowers. Preference toward funnel and stalk-disc flowers tended to be neutral (with one strong positive outlier --- *B. gerstaeckeri* --- for funnel flowers), reflecting the fact that these morphotypes tended to be rarely visited while also occurring at low abundance. Preference toward head flowers varied widely across bumble bee species, ranging from strongly positive (*B. psithyrus*, *B. soroeensis*) to strongly negative (e.g. *B. gerstaeckeri*, *B. mendax*).

@@

@@

Tongue length was a strong predictor of preference but not of observed visitation rates **(Figure 3)**. Preference toward funnel, bell, lip, flag, and head flowers exhibited significant GAM curves in response to bumble bee tongue length, and the model explained 77% of total deviance (adj. R^2^ = 0.728). For funnel and lip flowers, preference increased linearly with tongue length. Preference toward bell flowers and head flowers exhibited approximately quadratic responses with preference peaking among short-tongued species and declining sharply with increasing tongue length before leveling off. Preference toward flag flowers exhibited a unimodal peak among medium-tongued species. In contrast, observed visitation rates were predicted by tongue length only for head flowers, which exhibited a weak decline in preference from a peak among short-tongued species. The visitation rate model explained only 25% of total deviance (adj. R^2^ = 0.053). Again, these results did not differ when preference was based on the worker-only null model (Appendix 1).

@@

The relative abundance of the most visited floral morphotypes --- bell, lip, flag, and head flowers --- was strongly patterned by elevation and season **(Figure 5)**. Bell flowers peaked in relative abundance early in the season and at mid-elevation, and lip flowers peaked late in the season and at low elevation. Flag flowers peaked in mid-season and exhibited two distinct elevational peaks, the lower at around 1000 m.s.l and the higher at around 1800 m.s.l. Head flowers exhibited a mid-season peak at low elevation and a broad late season peak at around 1600 m.s.l. All smooths were significant (p << 0.0001), and the model explained 64.3% of total deviance (adj. R^2^ = 0.554).

@@

The morphotype preferences of bumble bee species were also strongly patterned through elevation and season. In some cases, though, patterns of preference mirrored patterns of floral availability, suggesting that bumble bees may exhibit functional responses [@Mysterud1998-rs] to floral abundance not entirely accounted for by null models that assume a linear proportionality between abundance and visitation. Preference for flag flowers was strongly associated with high elevation, except among the species that tended to avoid flag flowers (*B. pratorum* and *B. soroeensis*) **(Figure 1)**. It is also apparent that preference tends to be more dynamic where it is also higher overall. Bell flowers, which were positively selected overall by the short-tongued bumble bee species *B. pratorum*, *B. terrestris-lucorum*, and *B. soroeensis* **(Figure 1)**, also exhibited the most complex patterns of preference in elevation and time for these bumble bee species **(Figure 6)**. In contrast, the long and medium-tongued species *B. hortorum*, *B. pascuorum*, and *B. wurflenii*, which exhibited neutral or weakly positive overall selection toward bell flowers **(Figure 1)**, also show weak patterns of variation in preference toward bell flowers **(Figure 6)**. Nearly the reverse pattern is evident for flag flowers, reflecting their association with longer-tongued bumble bee species. Smooths were significant (p < 0.05) for all bee-morphotype pairs except hort-head, prat-flag, telu-head, and soro-flag, and the model explained 36.4% of total deviance (adj. R^2^ = 0.33).

@@

@@

Using ~13,000 bumble- bee-wildflower interactions, we characterized bumble bee floral selection in terms of generalized floral morphotypes and found evidence that bumble bee floral selection depends both on the directed process of morphotype preference and the neutral effect of floral abundance. Bumble bee species differed in their floral morphotype preferences, and these differences could be explained by trait-matching between floral morphotype and bumble bee tongue length. These preferences, however, varied by elevation and season, likely representing adaptive responses of bumble bee foraging behavior to elevational and seasonal shifts in floral resource composition.

@@

Our finding that bumble bee tongue length explained preference (i.e. null model residuals) much better than it explained observed visitation rates indicates that, in our study system, abundance and preference function jointly in bumble bee floral selection. This casts doubt on the *target use hypothesis* as a general rule of bumble bee foraging, though it is possible that it still obtains under certain circumstances [e.g. @Harmon-Threatt2017-bz]. In general, though, our findings support the hypothesis that bumble bee resource selection involves both the neutral effects of floral abundance and the directed effect of preference, meaning that analytical techniques designed to disentangle these effects, such as the null modeling method we employ, should be applied in studies of bumble bee foraging.

@@

We found that the most preferred morphotypes were bell, lip, and flag flowers, which is broadly consistent with visitation patterns reported by previous studies in central Europe [@Schneller2014-ry; @Neumayer1999-il; @Farnberger1996-xo], indicating that our findings are regionally generalizeable. These previous studies tended to find a stronger role of head flowers than indicated by our preference analysis, but this is probably because they characterized foraging in terms of simple visitation rates rather than null-model-inferred preference. We saw similarly high visitation rates toward head flowers, but our null-model analysis reveals that while head flowers were highly visited, they tended to be under-visited relative to their abundance.

@@

**Figure 1:** Observed visitation rates compared to null model simulations for each bumble bee species and floral morphotype (A). Gray bars depict 99% intervals of null model simulations and points depict observed visitation rates. Points are color-coded according to where they fall with respect to the 99% interval of the null model: red = above (positive preference), blue = below (negative preference), gray = within (nonsignificant preference). The difference between the observed value and the center of the null model interval can be interpreted as magnitude of preference. Each floral morphotype is illustrated (B) with a representative species common in our study system: 1 = *Potentilla erecta*, 2 = *Gentiana asclepiadea*, 3 = *Campanula glomerata*, 4 = *Silene nutans*, 5 = *Ajuga reptans*, 6 = *Lotus corniculatus*, 7 = *Centaurea jacea*. Illustrations are modifications of public domain works obtained from http://www.plantillustrations.org. For details of original sources, see Appendix 2.
